---
title: expand-grid
author: Matthew Routley
date: '2006-07-06'
slug: expand-grid
categories:
  - code
tags: []
---

Here&#8217;s a simple trick for creating experimental designs in R: use the function expand.grid.

A simple example is:

```{r eval=FALSE}
treatments <- LETTERS[1:4]
levels <- 1:3
experiment <- data.frame(expand.grid(treatment=treatments, 
                                     level=levels))
```

which produces:

```{r eval=FALSE}
treatment level
1          A     1
2          B     1
3          C     1
4          D     1
5          A     2
6          B     2
7          C     2
8          D     2
9          A     3
10         B     3
11         C     3
12         D     3
```

Now, if you want to randomize your experimental treatments, try:

```{r eval=FALSE}
experiment[sample(dim(experiment)[1]), ]
```

`sample` randomly chooses numbers from a vector the same length as the `experiment` data frame without replacement. The square brackets then use this random sample to subsample from the `experiment` data frame.